Hydras \& Co.: Formalized mathematics in Coq\\ for inspiration and entertainment

### 2022-04-23 doc/JFLA2022/Slides/Slides.tex
update slides about Gaia
changes to JFLA slides

\documentclass[10pt]{beamer}
\usetheme{Montpellier}

\usepackage{xspace, alltt}

@@

\newcommand{\canonseq}[2]{\mbox{$\{#1\}(#2)$}}

\title{Hydras \& Co.: Formalized mathematics in Coq\\
  for inspiration and entertainment
}
\date{JFLA, February 2022}
\author{
  Pierre Castéran \inst{1}
  \and
  Jérémy Damour \inst{2}
  \and
  Karl Palmskog \inst{3}
  \and Clément Pit-Claudel \inst{4}
  \and Théo Zimmermann \inst{5}
}

@@

\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{The \Hydras project}
  \begin{block}{}
    \begin{itemize}
    \item Experimental platform for the  development of \textcolor{lookcolor}{maintained} and \textcolor{lookcolor}{documented} libraries of formal proofs.
    \item A project of \community, developped on Github
      \begin{itemize}
      \item CI/CD
        \item Tools for collaboration (issues, PR, etc.)
      \end{itemize}
    \end{itemize}
  \end{block}
  \begin{block}{Contents}
    \begin{itemize}
    \item Evolved version  of two contributions: \textcolor{plugincolor}{Cantor} and \textcolor{plugincolor}{Additions}
    \item Maintenance of \textcolor{plugincolor}{Ackermann}
      (part of Russel O'Connor's \textcolor{plugincolor}{Goedel})
    \item Preliminary bridge to \textcolor{plugincolor}{Gaia} (José Grimm)
      \item Documentation in pdf (280p), generated  with \textcolor{lookcolor}{\alectr} and \LaTeX, continuously updated and deployed
    \end{itemize}
  \end{block}

@@

\begin{frame}
  \frametitle{What hydras are talking about?}
  \begin{block}{}
    \Hydras presents a consistent set of examples, which allow us to formalize 
    some \textcolor{mathcolor}{discrete math}, discuss
    \textcolor{coqstylecolor}{\coq style, specification and proof patterns}, and use\footnote{and depend on!} \textcolor{plugincolor}{a few plug-ins}.
  \end{block}
  \begin{block}{}
    \begin{description}
    \item[hydra battles:]

@@

    \caption{Three successive states of a hydra in a battle,
      at time $t=0,1,2$\label{life1}}
    \label{fig:round}
  \end{figure}

@@

    \end{tikzpicture}
    \caption{At $t=3,\dots,\textcolor{red}{x}-1,\textcolor{red}{x=?}$
      \label{life2}}
  \end{figure}
\end{frame}

@@

\begin{frame}
    \begin{block}{Mathematical formalization}

    \begin{itemize}
    \item   We associate a hydra to each ordinal \textcolor{mathcolor}{$\alpha<\epsilon_0$} (in Cantor normal form). For instance, the first hydra of our example is the image of \textcolor{mathcolor}{$\omega^{\omega^2+1}+1$}.
    \item A round (at time $t\in\mathbb{N}$) of a battle transforms an hydra associated with \textcolor{mathcolor}{$\alpha$} into the hydra associated with \textcolor{mathcolor}{$\canonseq{\alpha}{t+1}$}, the $t+1$-st element of the \emph{canonical sequence} of \textcolor{mathcolor}{$\alpha$}.
    \item We want to ``compute'' the length of sequences of the form 
      \textcolor{mathcolor}{$\langle \alpha_k=\alpha,\dots,\alpha_{i+1}=\canonseq{\alpha_i}{i+1},\dots, 1,0\rangle$}, where \textcolor{mathcolor}{$k\in\mathbb{N}$}.
      \item Such sequences are studied
      as \emph{$\alpha$-large sets of natural numbers}  in Ketonen \& Solovay's article
      \emph{Rapidly Growing {R}amsey Functions (1981)}.
    \end{itemize}
  \end{block}
\end{frame}

@@

  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Organization}
\begin{frame}
  \frametitle{Project  Organization}
  \begin{itemize}
  \item Integration with \community
  \item Documentation with \alectr
  \item Package dependencies
  \item Continuous integration
  \end{itemize}
\end{frame}

@@

%%%% TODO : Resize  !
{\scriptsize
\begin{table}[ht]
\centering
\scriptsize
\begin{tabular}{llllrr}
\hline
\textbf{Package name} &  \textbf{Version} & \textbf{Spec LOC} & \textbf{Proof LOC}\\
\hline
Pocklington & \texttt{8.12.0} & 825 & 3,798 \\
Goedel  & \texttt{8.13.0} & 2,799 & 10,762 \\
Gaia &  \texttt{1.12} & 28,850 & 124,839 \\
Hydra-battles & \texttt{0.5} & 14,368 & 48,385 \\
Addition-chains & \texttt{0.5} & 1,961 & 2,210 \\
Gaia-hydras &  \texttt{0.5} & 81 & 177 \\
\hline
\end{tabular}
\caption{\scriptsize Current numbers of lines of code for packages in the \Hydras galaxy.}
\label{tbl:loc}
\end{table}}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\documentclass[10pt]{beamer}
\usetheme{Montpellier}
\usepackage{xspace, alltt}
\usepackage[utf8x]{inputenc}

\usepackage{fontspec}
%\setmainfont{DejaVu Serif}

\setmonofont{DejaVu Sans Mono}[Scale=MatchLowercase]

@@

\newcommand{\canonseq}[2]{\mbox{$\{#1\}(#2)$}}
\newcommand{\rounds}{\mbox{\,\texttt{-+->}\,}}
\newcommand{\round}{\mbox{\,\texttt{-1->}\,}}
\newcommand{\rplus}[1]{\mbox{$\,\underset{#1}{\xrightarrow{\textcolor{white}{\,#1\,}}}\,$}}

\title{Hydras \& Co.: Formalized mathematics in Coq\\
  for inspiration and entertainment
}
\date{JFLA, June 2022}
\author{
  Pierre Castéran \inst{1}
  \and
  Jérémy Damour \inst{2}
  \and
  Karl Palmskog \inst{3}
  \and Clément Pit-Claudel \inst{4}
  \and Théo Zimmermann \inst{5}
}

@@

\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{The \Hydras project}
  \begin{block}{}
    \begin{itemize}
   \item Experimental platform for the  development of \textcolor{lookcolor}{maintained} and \textcolor{lookcolor}{documented} libraries of formal proofs.
    \item A project of \community, developped on Github
      \begin{itemize}
      \item CI/CD
        \item Tools for collaboration (issues, PR, etc.)
        \end{itemize}
           \item book in pdf (300p), generated  with \textcolor{lookcolor}{\alectr} and \LaTeX, continuously updated and deployed
    \end{itemize}
  \end{block}
  \begin{block}{Contents}
    \begin{itemize}
    \item Evolved version  of two contributions: \textcolor{plugincolor}{Cantor} and \textcolor{plugincolor}{Additions}
    \item Maintenance of \textcolor{plugincolor}{Ackermann}
      (part of Russel O'Connor's \textcolor{plugincolor}{Goedel})
    \item Bridge to \textcolor{plugincolor}{Gaia} (José Grimm) (\emph{in construction})

@@

\begin{frame}
  \frametitle{What hydras are talking about?}
  \begin{block}{}
    \Hydras presents a consistent set of examples, which allow us to formalize 
    some \textcolor{mathcolor}{discrete math}, discuss
    \textcolor{coqstylecolor}{\coq style, specification and proof patterns}, and depend on \textcolor{plugincolor}{a few plug-ins}.
  \end{block}
  \begin{block}{}
    \begin{description}
    \item[hydra battles:]

@@

    \caption{Three successive states of a hydra in a battle,
      at time $t=1,2,3$\label{life1}}
    \label{fig:round}
  \end{figure}

@@

    \end{tikzpicture}
    \caption{At $t=4,\dots,\textcolor{red}{x}-1,\textcolor{red}{x=?}$
      \label{life2}}
  \end{figure}
\end{frame}

@@

\begin{frame}
  \frametitle{The Ketonen-Solovay machinery\footnote{Rapidly Growing {R}amsey Functions (1981)}}
    \begin{block}{Formalization}

      $$\omega^{\omega^2+1}+1 \rplus{1} \omega^{\omega^2+1}
      \rplus{2} \omega^{\omega^2}\times 3 
      \rplus{3} \omega^{\omega^2}\times 2 +
      \omega ^ {\omega \times 4}, \dots,2,1,0$$

      \vspace{6pt}

      $$ \omega^{\omega^2+1}+1 \rplus{1,2,3,4,\dots,x} 0$$
    \end{block}

      \begin{block}{Small steps}
        At time $i$, 
           $\alpha \rplus{i} \canonseq{\alpha}{i}$,
        where $\canonseq{\alpha}{\_}$ is the \emph{canonical sequence} of $\alpha$.
        \begin{itemize}
        \item $\canonseq{0}{i}=0$
        \item $\canonseq{\beta+1}{i}=\beta$
         \item If $\lambda$ is a limit ordinal, then $\lambda$ is the $\omega$-limit of $\canonseq{\lambda}{\_}$.
        \end{itemize}
    \end{block}

      % We associate a hydra to each ordinal \textcolor{mathcolor}{$\alpha<\epsilon_0$} (in Cantor normal form). For instance, the first hydra of our example is the image of \textcolor{mathcolor}{$\omega^{\omega^2+1}+1$}.

    % \item We want to ``compute'' the length of sequences of the form 
    %   \textcolor{mathcolor}{$\langle \alpha_k=\alpha,\dots,\alpha_{i+1}=\canonseq{\alpha_i}{i+1},\dots, 1,0\rangle$}, where \textcolor{mathcolor}{$k\in\mathbb{N}$}.

\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}

     \begin{block}{}
       \begin{itemize}
       \item In Coq, canonical sequences are defined by structural recursion on ordinal terms in Cantor normal form.
         \input{../../movies/snippets/Canon/canonExamplesb}
       \item Not the same definition as in K. \& S.'s article, but same abstract properties (monotony, limits, \dots).
       \end{itemize}
     \end{block}
\end{frame}

@@

  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \frametitle{Hydras \& Gaia}
  \begin{block}{From \gaia to \Hydras}
    {\footnotesize
    \inputsnippets{./multA}}
  \end{block}
\end{frame}

\begin{frame}
  \begin{block}{From \Hydras to \gaia}
    {\footnotesize
      \inputsnippets{./canonDef}
      \inputsnippets{./gcanonLimitOf}}
  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Organization}
\begin{frame}
  \frametitle{Project  Organization}
  \begin{itemize}
  \item Integration with \community
  \item Documentation with \alectr
  \item Package dependencies
  \item Continuous integration
  \end{itemize}
\end{frame}

@@

%%%% TODO : Resize  !
{\scriptsize
\begin{table}[ht]
\centering
\scriptsize
\begin{tabular}{llllrr}
\hline
\textbf{Package name} &  \textbf{Version} & \textbf{Spec LOC} & \textbf{Proof LOC}\\
\hline
Pocklington & \texttt{8.12.0} & 825 & 3,798 \\
Goedel  & \texttt{8.13.0} & 2,799 & 10,762 \\
Gaia &  \texttt{1.12} & 28,850 & 124,839 \\
Hydra-battles & \texttt{0.5} & 14,368 & 48,385 \\
Addition-chains & \texttt{0.5} & 1,961 & 2,210 \\
Gaia-hydras &  \texttt{0.8} & 677 & 2,023 \\
\hline
\end{tabular}
\caption{\scriptsize Current numbers of lines of code for packages in the \Hydras galaxy.}
\label{tbl:loc}
\end{table}}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
